% ------------------------------------------------------------
\chapter{Implementation}
% ------------------------------------------------------------
 
\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}
 
Chapter~3 established the architectural design of the system —
defining what each component should do and how the layers relate to
one another. This chapter documents how that design was realised in
practice. The focus shifts from \emph{what was intended} to
\emph{what was built}: the specific hardware, software tools,
libraries, and engineering decisions that translate the architectural
model into a functioning prototype.
 
The implementation encompasses the following primary components:
 
\begin{itemize}
    \item The development environment, including hardware platform
          selection, operating system, programming tools, and
          dependency management.
    \item The embedded inference pipeline, covering model format
          conversion and the deployment challenges encountered on
          the target hardware.
    \item The Raspberry Pi software stack, including the multi-threaded
          service architecture, the Finite State Machine runtime, and
          the camera acquisition pipeline.
    \item The Arduino firmware, covering the command processor,
          actuator abstraction, and failsafe mechanism.
    \item The communication layer, including the WebSocket client
          implementation and the UART serial protocol.
\end{itemize}
 
A recurring theme throughout this chapter is the gap between
theoretical deployment assumptions and practical hardware constraints.
Several implementation decisions — most notably the choice of
inference runtime and model format — were driven not by design
preference but by the limitations of the available embedded hardware
and the compatibility boundaries of the available software toolchains.
These decisions are documented candidly so that future work can build
on the lessons learned.
 
% ------------------------------------------------------------
\section{Development Environment}
\label{sec:devenv}
% ------------------------------------------------------------
 
This section documents the hardware platform, software environment,
and dependency management process used throughout the implementation
phase. Establishing a reproducible and stable development environment
was itself a significant engineering effort, particularly with respect
to AI inference dependencies on constrained embedded hardware — a
challenge that ultimately influenced core architectural decisions and
is examined in detail in Section~\ref{sec:dep_challenges}.
 
% ............................................................
\subsection{Hardware Platform}
% ............................................................
 
The embedded hardware used in this project consists of three physical
components, each serving a distinct role in the prototype system.
 
\paragraph{Raspberry Pi 3 Model B.}
The Raspberry Pi 3B+ is the central processing unit of the embedded
system. It features a 1.4\,GHz quad-core ARM Cortex-A53 processor,
1\,GB of LPDDR2 RAM, integrated 802.11ac dual-band Wi-Fi, and a
dedicated CSI camera port~\cite{upton2016raspberry}. While the
Cortex-A53 is a capable general-purpose processor, it lacks a
dedicated neural processing unit (NPU) or hardware floating-point
accelerator optimised for inference workloads — a limitation that
proved consequential during the model deployment phase (see
Section~\ref{sec:dep_challenges}).
 
\paragraph{Arduino Uno.}
The Arduino Uno is based on the ATmega328P 8-bit microcontroller
running at 16\,MHz with 2\,KB of SRAM and 32\,KB of flash
storage~\cite{monk2016electronics}. It is responsible exclusively for
low-level actuator control: receiving UART commands from the Raspberry
Pi and generating the PWM signals required by motor drivers and relay
circuits. Its limited resources are well-matched to this role — the hi
firmware is small and its execution timing is deterministic.
 
\paragraph{Raspberry Pi Camera Module 3.}
The Camera Module 3 connects to the Raspberry Pi via the CSI ribbon
interface and provides up to 12\,MP resolution through a Sony IMX708
sensor~\cite{picamera2023}. For inference purposes frames are captured
at a reduced resolution to balance image quality against the
computational cost of preprocessing and model inference on the
embedded processor.
 
% ............................................................
\subsection{Software Environment}
% ............................................................
 
\paragraph{Raspbian GNU/Linux 13 (Trixie).}
The operating system is Raspbian (Raspberry Pi OS) based on Debian 13
``Trixie'', a 64-bit Linux distribution maintained by the Raspberry Pi
Foundation and optimised for the Raspberry Pi hardware
family~\cite{raspios2024}. The 64-bit userspace is important for AI
inference libraries, many of which distribute pre-built wheels only
for \texttt{aarch64} and not for the legacy 32-bit \texttt{armv7l}
architecture.
 
\paragraph{Python 3.13.5.}
Python 3.13.5 is the primary application programming language for all
Raspberry Pi software. Python was chosen for its extensive ecosystem
of computer vision and machine learning libraries, its compatibility
with the inference runtimes under consideration, and its suitability
for rapid prototyping of complex, multi-threaded
systems~\cite{python313}.
 
\paragraph{Arduino IDE.}
The Arduino Integrated Development Environment is used to write,
compile, and flash the C++ firmware onto the Arduino Uno. The IDE
provides the AVR-GCC toolchain and the \texttt{avrdude} programmer
required to deploy firmware over USB without additional
tools~\cite{arduinoide}.
 
\paragraph{Raspberry Pi Connect.}
Raspberry Pi Connect is a secure remote access service provided by
the Raspberry Pi Foundation that enables shell and desktop access to
a Raspberry Pi over the internet without configuring port forwarding
or a VPN~\cite{rpconnect2024}. It was used during development to
remotely monitor, debug, and redeploy software on the embedded device
without requiring physical access.
 
\paragraph{OpenCV.}
OpenCV (Open Source Computer Vision Library) is an open-source
library providing optimised implementations of image processing
algorithms~\cite{bradski2000opencv}. In this project it is used for
frame preprocessing (resizing, colour space conversion, normalisation)
before frames are passed to the inference pipeline.
 
\paragraph{ONNX Runtime.}
ONNX Runtime is a cross-platform inference engine for models in the
Open Neural Network Exchange format~\cite{onnxruntime2023}. It was
selected as the primary inference backend for the server-side
deployment path due to its broad operator support, active maintenance,
and strong performance on Linux x86-64 hardware.
 
\paragraph{TensorFlow Lite.}
TensorFlow Lite is Google's inference framework for mobile and
embedded devices~\cite{david2021tensorflow}. It was evaluated as an
alternative embedded inference backend during the dependency
investigation described in Section~\ref{sec:dep_challenges}.
 
% ............................................................
\subsection{Build and Dependency Management}
% ............................................................
 
Python dependencies for the Raspberry Pi software stack are managed
using a virtual environment (\texttt{venv}) to isolate project
packages from the system Python installation. CMake is used where
native extension modules require compilation from source — a
requirement that arose during several inference library installation
attempts on the ARM platform.
 
\subsubsection{Dependency and Deployment Challenges}
\label{sec:dep_challenges}
 
One of the most time-consuming aspects of the implementation phase was
resolving inference runtime dependencies on the Raspberry Pi 3B+.
This section documents the investigation process and the decisions it
produced, because these constraints directly shaped the final
deployment architecture.
 
\paragraph{Stage 1 — Direct RF-DETR Inference.}
The initial intention was to run the RF-DETR Nano model directly on
the Raspberry Pi using its native PyTorch weights. Profiling quickly
revealed that the model was too computationally demanding for the
Cortex-A53 processor: inference times were far outside the range
acceptable for real-time operation, and peak memory usage approached
the device's 1\,GB RAM limit. A lighter-weight, hardware-optimised
format was required.
 
\paragraph{Stage 2 — ONNX Runtime on the Raspberry Pi.}
The model was exported to ONNX format (as described in
Section~\ref{sec:quantization}) and ONNX Runtime was identified as
the preferred embedded inference engine due to its Linux compatibility
and active ARM support. However, after export, no pre-built
\texttt{.whl} package for ONNX Runtime was available that matched
the combination of Python 3.13.5 and the \texttt{aarch64} architecture
of the Raspberry Pi 3B+. Attempts to install wheels built for adjacent
Python versions (3.11, 3.12) failed due to ABI incompatibilities.
Compiling ONNX Runtime from source requires CMake, a C++17-capable
compiler, and a build process that exceeds the memory capacity of the
device itself.
 
Downgrading the Python runtime to a version with available ONNX
Runtime wheels was explored, but this introduced conflicts with other
project dependencies that required Python~3.13 APIs, making this path
unviable.
 
\paragraph{Stage 3 — TensorFlow Lite via ONNX Conversion.}
TensorFlow Lite was considered as an alternative embedded runtime,
as pre-built TFLite interpreter packages are available for a wider
range of ARM Python configurations. Because RF-DETR has no native
TFLite export path, a two-stage conversion pipeline was required:
 
\begin{center}
RF-DETR (PyTorch) $\xrightarrow{\text{torch.onnx.export}}$
ONNX $\xrightarrow{\text{onnx2tf / ai\_edge\_torch}}$ TFLite
\end{center}
 
A compatible Python runtime version was identified under which the
TFLite interpreter package installed successfully. However, this
runtime version was incompatible with the system-installed OpenCV
and \texttt{picamera2} bindings, which are compiled against the
system Python and cannot be imported from a different interpreter.
Custom wrapper solutions were prototyped to bridge frame acquisition
from \texttt{picamera2} into the alternative runtime environment,
but these introduced additional fragility and maintenance overhead.
 
After completing the conversion pipeline, the resulting
\texttt{.tflite} model failed to load on the target runtime. The
RF-DETR architecture relies on custom attention and decoder operations
that are not part of the standard TFLite operator set; the conversion
tools emitted these as \emph{unsupported operators}, which the TFLite
interpreter rejected at model initialisation time. Multiple conversion
configurations and operator compatibility flags were tested without
success.
 
\paragraph{Resolution — Remote Inference Fallback.}
Having exhausted the available embedded runtime options for the current
hardware, the decision was made to adopt the remote inference fallback
path described in Section~\ref{sec:inference_pipeline}: frames are
streamed from the Raspberry Pi to the backend server over WebSocket,
where ONNX Runtime on the x86-64 server executes inference and returns
detection results to the Pi. This path has no operator-support
constraints and imposes no memory or compute requirements on the
embedded device itself.
 
This outcome reinforces the value of designing the inference pipeline
with a fallback mode from the outset (Section~\ref{sec:inference_pipeline}):
the architectural decision to separate the inference interface from its
implementation meant that switching from local to remote execution
required no changes to the surrounding software stack.
 

\section{AI Model Deployment and Evaluation}
\label{sec:model_deployment}

This section addresses one of the central challenges of the project:
deploying an RF-DETR Nano object detection model onto a
resource-constrained embedded device. The deployment process involved
evaluating multiple model formats, executing a multi-stage conversion
pipeline, and navigating a series of hardware and runtime
compatibility issues that ultimately determined the final system
architecture. Because these decisions directly affect the system's
ability to fulfil its core purpose — real-time pest detection in the
field — this section is among the most significant in the chapter.

% ............................................................
\subsection{Original Detection Model}
\label{sec:original_model}
% ............................................................

The object detection model used in this project was inherited from
Phase~1 of the work, in which an RF-DETR Nano model was trained on a
labelled agricultural pest dataset. RF-DETR Nano is a transformer-based
real-time detector built on a DINOv2 vision transformer backbone and a
DETR-style encoder–decoder head, as described in
Section~\ref{sec:rfdetr}. The model was delivered in its native
Roboflow RF-DETR checkpoint format — a PyTorch \texttt{.pt} weight
file paired with the \texttt{rfdetr} Python package for inference.

While this format is convenient for experimentation and retraining on
a workstation, it is not suitable for direct embedded deployment for
two reasons. First, PyTorch carries a large runtime footprint that
exceeds the available memory of the Raspberry Pi 3B+ during model
loading. Second, the PyTorch inference path does not take advantage of
hardware-level optimisations available through dedicated inference
runtimes. Format conversion was therefore a necessary precondition for
any embedded deployment attempt.

% ............................................................
\subsection{Quantization and Precision Formats}
\label{sec:quant_formats}
% ............................................................

Quantization reduces the numerical precision of a model's weights and
activations, lowering memory consumption and accelerating inference on
hardware that handles low-precision arithmetic efficiently. Four
precision configurations were considered during this project, each
representing a different point on the accuracy-versus-efficiency
trade-off curve (quantization theory was introduced in
Section~\ref{sec:quantization}; this section focuses on their
practical implications for deployment).

\paragraph{FP32 — Baseline.}
Full 32-bit IEEE~754 floating-point is the default training precision.
Every weight and activation is stored as a 4-byte value, giving the
widest dynamic range and the highest numerical fidelity. On the
Raspberry Pi 3B+'s Cortex-A53, FP32 arithmetic is supported in
hardware via the ARM NEON SIMD unit, but without a dedicated neural
accelerator each multiply-accumulate operation still occupies a general
purpose pipeline stage. FP32 was used as the accuracy baseline against
which all quantized variants were compared.

\paragraph{FP16 — Half Precision.}
FP16 halves the storage requirement of each parameter (2~bytes) and
can double effective memory bandwidth. On server-class GPUs with
native FP16 tensor cores it also halves arithmetic latency. However,
the standard \texttt{tflite\_runtime} distributed for ARM CPUs does
\emph{not} include FP16 CPU kernels. Attempting to run an FP16 TFLite
model on the Raspberry Pi produces runtime errors for operators such
as \texttt{FlexErf} and \texttt{DEQUANTIZE}, which are absent from
the ARM build of the interpreter. FP16 was therefore excluded from
the embedded deployment path despite its theoretical appeal.

\paragraph{INT8 — Static Integer Quantization.}
INT8 maps each floating-point value to a signed 8-bit integer in
$[-128,\,127]$ using a per-layer \emph{scale} and \emph{zero-point}
derived from a calibration dataset. The result is a $4\times$ reduction
in model size compared to FP32 and significantly faster inference on
processors with integer multiply-accumulate units. Full static INT8
quantization requires a representative calibration set and full
TensorFlow (not just \texttt{tflite\_runtime}) to run the
\texttt{TFLiteConverter} pipeline — a dependency that proved
problematic on the embedded device.

\paragraph{Dynamic Range INT8 — Post-Training Quantization.}
Dynamic range quantization quantizes model \emph{weights} to INT8
offline but leaves \emph{activations} in floating-point, quantizing
them dynamically at inference time. No calibration dataset is required,
making it the simplest form of quantization to apply. It was evaluated
as the most accessible path to a smaller model footprint while
minimising conversion complexity. The \texttt{tf.lite.Optimize.DEFAULT}
flag in \texttt{TFLiteConverter} enables this mode.

% ............................................................
\subsection{Neural Network Operators (Ops)}
\label{sec:ops}
% ............................................................

A neural network is represented internally as a \emph{computational
graph}: a directed acyclic graph (DAG) in which each node is an
\textbf{operator} (op) and the edges are tensors flowing between
them~\cite{abadi2016tensorflow}. Every mathematical operation in the
network — a 2-D convolution, a matrix multiplication, a softmax
normalisation, a layer normalisation, a GELU activation — corresponds
to one or more op nodes in this graph.

An inference runtime executes a model by traversing this graph and
dispatching each op to a hardware-specific \emph{kernel}
implementation. A runtime only supports the ops for which it provides
kernels. If a model's graph contains an op that has no kernel in the
target runtime, the runtime cannot execute the model and raises an
\emph{unsupported operator} error at load time.

This constraint is particularly acute for transformer-based models
such as RF-DETR Nano. Unlike CNN-based detectors whose graphs consist
almost entirely of convolution and pooling ops (which every major
runtime supports), RF-DETR's DINOv2 backbone and DETR decoder rely on
operations including:

\begin{itemize}
    \item Multi-head self-attention and cross-attention
    \item Layer normalisation with learned affine parameters
    \item GELU (Gaussian Error Linear Unit) activation
    \item Custom position encoding operations
    \item Deformable attention (in the DETR decoder)
\end{itemize}

Several of these operations are not part of the \emph{standard}
TFLite operator set (the \texttt{TFLITE\_BUILTINS} kernel library).
They exist in TFLite only as \textbf{Flex ops} (also called Select
ops) — a mechanism that embeds a subset of the full TensorFlow runtime
inside the TFLite interpreter to handle non-standard
ops~\cite{david2021tensorflow}. However, the Flex delegate requires
linking against a much larger library (\texttt{libtensorflow.so}),
which is not included in the standard \texttt{tflite\_runtime} wheel
distributed for ARM devices. This mismatch between the ops required by
RF-DETR and the ops available in the lightweight ARM TFLite build is
the fundamental technical blocker described in
Section~\ref{sec:dep_challenges_detail}.

% ............................................................
\subsection{Model Conversion Pipeline}
\label{sec:conversion_pipeline}
% ............................................................

The conversion pipeline proceeded in two stages: first from the native
RF-DETR PyTorch checkpoint to ONNX, and then from ONNX to TFLite.

\subsubsection{Stage 1 — RF-DETR to ONNX}

Export from RF-DETR to ONNX is natively supported by the Roboflow
\texttt{rfdetr} Python package. The model's \texttt{export()} method
invokes \texttt{torch.onnx.export} internally, tracing the PyTorch
computation graph and serialising it to a self-contained
\texttt{.onnx} file. The export process is straightforward and
requires only the pretrained weights path and the number of detection
classes:

\begin{lstlisting}[language=Python,
                   caption={RF-DETR Nano export to ONNX format.},
                   label={lst:rfdetr_export}]
from rfdetr import RFDETRNano

model = RFDETRNano(
    pretrain_weights="path/to/weights.pt",
    num_classes=NUM_CLASSES
)

model.export(output_dir="path/to/models")
\end{lstlisting}

The resulting \texttt{.onnx} file contains the full computational
graph with FP32 weights and is immediately usable with ONNX Runtime
on x86-64 hardware.

\subsubsection{Stage 2 — ONNX to TFLite}

Because RF-DETR provides no direct TFLite export path, conversion from
ONNX to TFLite was performed using the \texttt{onnx2tf} library, which
translates an ONNX graph into a TensorFlow SavedModel and then invokes
\texttt{TFLiteConverter} to produce the final \texttt{.tflite} file.

Two \texttt{onnx2tf} options required careful configuration to avoid
generating models with unsupported operators on the target runtime:

\begin{lstlisting}[language=Python,
                   caption={ONNX to TFLite conversion using onnx2tf.},
                   label={lst:onnx2tf}]
import onnx2tf

onnx2tf.convert(
    input_onnx_file_path=onnx_path,
    output_folder_path=output_dir,
    output_integer_quantized_tflite=False,
    output_float16_quant_tflite=False
)
\end{lstlisting}

\begin{itemize}
    \item \texttt{output\_integer\_quantized\_tflite=False} — Prevents
          \texttt{onnx2tf} from applying its own integer quantization
          pass, which can introduce operator patterns that require the
          Flex delegate. Disabling this flag produces a clean FP32
          graph that can be quantized separately in a controlled
          manner.

    \item \texttt{output\_float16\_quant\_tflite=False} — Prevents
          generation of an FP16 model. As established in
          Section~\ref{sec:quant_formats}, the standard
          \texttt{tflite\_runtime} for ARM does not include FP16 CPU
          kernels; an FP16 model would fail at runtime with
          \texttt{FlexErf} and \texttt{DEQUANTIZE} errors. This flag
          is critical for producing a model that can be executed
          without the Flex delegate.
\end{itemize}

An optional dynamic-range INT8 quantization step was applied after
the initial FP32 conversion using TensorFlow's \texttt{TFLiteConverter}
directly:

\begin{lstlisting}[language=Python,
                   caption={Optional dynamic-range INT8 quantization
                            via TFLiteConverter.},
                   label={lst:int8_quant}]
import tensorflow as tf

converter = tf.lite.TFLiteConverter.from_saved_model(saved_model_dir)
converter.optimizations = [tf.lite.Optimize.DEFAULT]
converter.target_spec.supported_ops = [tf.lite.OpsSet.TFLITE_BUILTINS]
tflite_model = converter.convert()
\end{lstlisting}

Setting \texttt{supported\_ops} to \texttt{TFLITE\_BUILTINS} alone is
the key constraint: it explicitly forbids the converter from emitting
Flex ops for any operation that cannot be expressed in the built-in
operator set. If a subgraph cannot be represented without Flex ops,
the converter raises an error rather than silently producing a model
that will fail on a standard \texttt{tflite\_runtime}. This flag
therefore functions as an early compatibility check.

% ............................................................
\subsection{Deployment Challenges and Resolution}
\label{sec:dep_challenges_detail}
% ............................................................

Despite the careful operator configuration applied during conversion,
the resulting TFLite models failed to load on the Raspberry Pi 3B+.
This section documents the specific failure modes encountered and the
reasoning behind the final architectural decision.

\subsubsection{Unsupported Operator Errors}

The core issue is that RF-DETR Nano's computational graph, even after
conversion, contains operator patterns that fall outside the
\texttt{TFLITE\_BUILTINS} operator set. The DETR decoder's
deformable attention mechanism and the DINOv2 backbone's layer
normalisation variants are expressed, after ONNX graph tracing, as
composite subgraphs whose constituent ops are not individually
supported by the ARM \texttt{tflite\_runtime} kernel library.

The \texttt{TFLITE\_BUILTINS} constraint in Listing~\ref{lst:int8_quant}
caused \texttt{TFLiteConverter} to raise conversion errors for these
subgraphs, confirming the incompatibility before runtime. Relaxing
the constraint to permit Flex ops (\texttt{SELECT\_TF\_OPS}) produced
a loadable \texttt{.tflite} file, but that file requires the Flex
delegate library (\texttt{libtensorflow\_lite\_c.so}), which is not
bundled with any \texttt{tflite\_runtime} wheel available for
Python on \texttt{aarch64}.

\subsubsection{Python Runtime Compatibility Conflict}

A \texttt{tflite\_runtime} wheel compatible with the device's
\texttt{aarch64} architecture was identified for an older Python
version (3.11). Switching to this runtime resolved the TFLite
installation issue but introduced a conflict with system-level
packages: the \texttt{picamera2} library and the system OpenCV
installation are compiled against the system Python (3.13) and cannot
be imported from a different interpreter. Custom bridging code was
prototyped to transfer frames between the two runtimes via shared
memory, but this approach added fragility and latency that were
inconsistent with the real-time requirements of the system.

\subsubsection{Resource Constraints}

Beyond operator compatibility, the Raspberry Pi 3B+ imposes hard
resource constraints that any embedded inference solution must satisfy:

\begin{itemize}
    \item \textbf{Memory.} The RF-DETR Nano model in FP32 requires
          approximately 100--200\,MB for weights alone; combined with
          the operating system, the camera pipeline, and the
          communication stack, total RAM usage approaches the 1\,GB
          ceiling of the device.

    \item \textbf{Compute.} The attention mechanism in the DETR decoder
          has quadratic complexity with respect to the number of
          spatial tokens, making it significantly more expensive per
          frame than a comparable CNN-based detector on a processor
          without matrix multiplication hardware.
\end{itemize}

\subsubsection{Resolution — Remote Inference Architecture}

After exhausting the available embedded runtime options, the decision
was made to adopt the remote inference fallback path that had been
built into the architecture from the outset
(Section~\ref{sec:inference_pipeline}). Under this configuration,
the Raspberry Pi streams compressed frames to the backend server over
the WebSocket channel; ONNX Runtime on the server's x86-64 CPU
executes inference and returns structured detection results
(bounding boxes, class labels, confidence scores) to the Pi.

This resolution carries no operator-support constraints — ONNX Runtime
on x86-64 supports the complete RF-DETR operator set natively —
and places no additional compute or memory burden on the embedded
device. The Raspberry Pi's role is reduced to frame acquisition,
compression, and transmission, tasks that are well within its
capabilities.

Critically, the architectural separation between the inference
interface and its implementation (Section~\ref{sec:software}) meant
that this change required no modification to the surrounding software
stack: the Inference Service simply selects the remote execution
provider at startup rather than a local one.

Table~\ref{tab:deployment_stages} provides a consolidated summary of
all deployment approaches evaluated, their specific failure modes,
and their outcomes.

\begin{table}[h]
\centering
\caption{Summary of model deployment approaches and outcomes.}
\label{tab:deployment_stages}
\begin{tabular}{p{3.2cm} p{4.2cm} p{4.2cm}}
\hline
\textbf{Approach} & \textbf{Failure Mode} & \textbf{Outcome} \\
\hline
PyTorch inference (native RF-DETR) &
    Memory and compute exceed Pi 3B+ capacity &
    Abandoned \\
ONNX Runtime on Pi &
    No compatible \texttt{aarch64} wheel for Python 3.13; source build exceeds device RAM &
    Abandoned \\
TFLite FP32 (TFLITE\_BUILTINS) &
    RF-DETR ops outside built-in set; converter raises unsupported op error &
    Abandoned \\
TFLite with Flex delegate &
    Flex delegate library absent from ARM \texttt{tflite\_runtime} wheel &
    Abandoned \\
TFLite on Python 3.11 runtime &
    Conflicts with system \texttt{picamera2} and OpenCV bindings &
    Abandoned \\
Remote inference via ONNX Runtime (server) &
    None &
    \textbf{Adopted} \\
\hline
\end{tabular}
\end{table}

% ============================================================
%  Chapter 4 – Section 4.4
%  Arduino Firmware Implementation
% ============================================================

% ------------------------------------------------------------
\section{Arduino Firmware Implementation}
\label{sec:arduino_firmware}
% ------------------------------------------------------------

This section documents the implementation of the Arduino Uno firmware
that controls the robot's actuation hardware. As established in the
architectural design (Section~\ref{sec:software}), the Arduino operates
as a serial-controlled actuator node: it accepts high-level text
commands from the Raspberry Pi, translates them into hardware signals,
and enforces safety constraints autonomously without relying on the
host system. The firmware is written in C++ using the Arduino framework
and is structured into four functional modules that mirror the
architecture defined in Section~\ref{sec:software}.

% ............................................................
\subsection{Firmware Structure}
\label{sec:firmware_structure}
% ............................................................

The firmware manages three motor driver circuits controlling a total
of five motors: four DC motors for differential drive mobility (two
per driver board) and one pump motor for pesticide delivery. The
physical pin assignment is fixed at compile time and documented in
Table~\ref{tab:pin_map}.

\begin{table}[h]
\centering
\caption{Arduino Uno pin assignment for motor and pump control.}
\label{tab:pin_map}
\begin{tabular}{llll}
\hline
\textbf{Motor} & \textbf{Driver} & \textbf{Speed (PWM)} & \textbf{Direction Pins} \\
\hline
Motor 0 & Driver 1 & Pin 3  & A0 (IN1), A1 (IN2) \\
Motor 1 & Driver 1 & Pin 5  & A2 (IN3), A3 (IN4) \\
Motor 2 & Driver 2 & Pin 6  & A4 (IN1), A5 (IN2) \\
Motor 3 & Driver 2 & Pin 9  & Pin 2 (IN3), Pin 4 (IN4) \\
Pump    & Driver 3 & Pin 10 & Pin 7 (direction) \\
\hline
\end{tabular}
\end{table}

Analogue pins (A0--A5) are configured as digital outputs using
\texttt{pinMode}, a standard Arduino technique that reclaims analogue
input pins for general digital I/O when the analogue-to-digital
converter is not required~\cite{monk2016electronics}.

The firmware is organised into four modules, each implemented as a
logical code section with clearly delineated responsibilities:

\begin{itemize}
    \item \textbf{Pin Initialisation} — configures all output pins
          and sets every actuator to a safe idle state at startup.
    \item \textbf{Actuator Control Layer} — provides \texttt{setMotor}
          and \texttt{setPump} functions that encapsulate all
          hardware signal generation.
    \item \textbf{Command Processor} — reads, parses, and dispatches
          newline-terminated ASCII commands received over UART.
    \item \textbf{Failsafe Monitor} — checks for communication
          timeout and triggers an emergency stop when necessary.
\end{itemize}

The \texttt{setup()} function initialises the serial port at 9600~baud,
calls \texttt{initializePins()}, invokes \texttt{safeStop()} to
guarantee a known-idle hardware state, and records the startup
timestamp for the failsafe clock. The \texttt{loop()} function runs
two tasks on every iteration: it services incoming serial bytes
(non-blocking) and calls \texttt{checkFailsafe()} to evaluate the
communication timeout condition.

\bigskip
\noindent\textbf{[DIAGRAM 1 — Firmware Module Diagram]}\\
\textit{Four stacked boxes: Command Processor (top, receives UART
bytes from left), Actuator Control Layer (middle, drives PWM/GPIO
signals to the right toward motor driver ICs), Pin Initialisation
(bottom, runs once at startup). The Failsafe Monitor sits as a side
module with a monitoring arrow pointing into the Actuator Control
Layer. Annotate the UART input with baud rate (9600) and newline
delimiter. Place after this subsection.}
\bigskip

% ............................................................
\subsection{Serial Communication Protocol}
\label{sec:serial_protocol}
% ............................................................

Communication between the Raspberry Pi and the Arduino uses a
lightweight, human-readable ASCII protocol over UART at 9600~baud.
Commands are plain text strings terminated by a newline character
(\texttt{\textbackslash n}); carriage return characters
(\texttt{\textbackslash r}) are silently discarded to ensure
compatibility with both Unix and Windows line endings. Each command
fits within a 64-byte buffer, a constraint that bounds memory usage
on the ATmega328P's 2\,KB SRAM~\cite{atmel2015atmega328p}.

\subsubsection{Command Format}

The firmware supports six commands, summarised in
Table~\ref{tab:commands}. Commands are matched case-insensitively
using \texttt{strcasecmp} and \texttt{strncasecmp}, making the
protocol tolerant of capitalisation variations from the host.

\begin{table}[h]
\centering
\caption{Arduino serial command set.}
\label{tab:commands}
\begin{tabular}{p{4.5cm} p{3cm} p{4.5cm}}
\hline
\textbf{Command} & \textbf{Parameters} & \textbf{Description} \\
\hline
\texttt{PING}                       & None                         & Connectivity check \\
\texttt{CHECK}                      & None                         & Motor subsystem health check \\
\texttt{STOP\_ALL} / \texttt{STOP}  & None                         & Emergency stop all actuators \\
\texttt{GET\_STATUS}                & None                         & Report all actuator states \\
\texttt{SET\_MOTOR <id> <spd> <dir>} & id: 0--3 or P; spd: 0--255; dir: 0/1 & Set motor or pump speed and direction \\
\texttt{SET\_PUMP <spd> <dir>}      & spd: 0--255; dir: 0/1        & Set pump speed and direction \\
\hline
\end{tabular}
\end{table}

\noindent The \texttt{SET\_MOTOR} command accepts a motor identifier
of \texttt{0}--\texttt{3} for the four drive motors or \texttt{P}
for the pump, unifying motor and pump control under a single command
verb. Speed is an unsigned 8-bit integer (0--255) mapped directly to
the PWM duty cycle via \texttt{analogWrite}; direction is a binary
flag where \texttt{1} denotes forward and \texttt{0} denotes
reverse~\cite{monk2016electronics}.

Any command that cannot be parsed is treated as an unknown command:
\texttt{safeStop()} is called immediately and an
\texttt{ERROR:UNKNOWN\_COMMAND} response is returned. This
fail-on-unknown design ensures that a corrupted or malformed message
does not leave actuators in an undefined state~\cite{kopetz2011real}.

\subsubsection{Response Format}

The firmware always responds to every command with a newline-terminated
ASCII string, enabling the Raspberry Pi's Communication Service to
detect missing acknowledgements within a timeout window. Four response
types are defined:

\begin{itemize}
    \item \texttt{PONG} — response to \texttt{PING}; confirms
          the serial link is alive.
    \item \texttt{MOTORS\_OK} — response to \texttt{CHECK}; confirms
          the firmware is initialised and ready.
    \item \texttt{OK} — confirms successful execution of a
          \texttt{SET\_MOTOR}, \texttt{SET\_PUMP}, or
          \texttt{STOP\_ALL} command.
    \item \texttt{STATUS:<m0spd>,<m0dir>,...,<pumpspd>,<pumpdir>}
          — a comma-separated string reporting the speed and direction
          of all five actuators in response to \texttt{GET\_STATUS}.
    \item \texttt{ERROR:<reason>} — reports a validation failure or
          safety event. Defined error codes are listed in
          Table~\ref{tab:errors}.
\end{itemize}

\begin{table}[h]
\centering
\caption{Firmware error response codes.}
\label{tab:errors}
\begin{tabular}{ll}
\hline
\textbf{Error Code} & \textbf{Cause} \\
\hline
\texttt{ERROR:UNKNOWN\_COMMAND}      & Command not recognised \\
\texttt{ERROR:INVALID\_MOTOR\_ID}   & Motor ID outside range 0--3 \\
\texttt{ERROR:INVALID\_MOTOR\_PARAMS} & Malformed speed/direction arguments \\
\texttt{ERROR:INVALID\_PUMP\_PARAMS} & Malformed pump arguments \\
\texttt{ERROR:CMD\_TOO\_LONG}        & Command exceeded 64-byte buffer \\
\texttt{ERROR:FAILSAFE\_STOP}        & Communication timeout; actuators halted \\
\hline
\end{tabular}
\end{table}

% ............................................................
\subsection{Actuator Control Layer}
\label{sec:actuator_layer}
% ............................................................

The Actuator Control Layer provides two functions — \texttt{setMotor}
and \texttt{setPump} — that form the only interface through which the
rest of the firmware interacts with hardware. All direct pin
manipulation is confined to these functions; no other module writes to
PWM or GPIO registers.

\subsubsection{Motor Control}

Each DC motor is driven by an H-bridge motor driver IC. An H-bridge
allows current to flow in either direction through the motor by
selecting the appropriate transistor pair, thereby controlling rotation
direction independently of speed~\cite{horowitz2015art}. The Arduino
controls an H-bridge through three signals per motor:

\begin{itemize}
    \item \textbf{Two direction pins (IN1/IN2 or IN3/IN4).}
          Setting IN1 HIGH and IN2 LOW drives the motor forward;
          reversing the levels drives it backward. Setting both LOW
          applies a passive brake.
    \item \textbf{One PWM speed pin (ENA/ENB).}
          \texttt{analogWrite(pin, speed)} outputs a PWM signal
          whose duty cycle is proportional to \texttt{speed}/255,
          regulating the average voltage — and therefore the torque
          and rotational velocity — applied to the
          motor~\cite{monk2016electronics}.
\end{itemize}

The \texttt{setMotor} function validates the motor identifier
(rejecting values outside 0--3), clamps the speed to the range
$[0, 255]$ using Arduino's \texttt{constrain()}, normalises the
direction to a binary value, updates the corresponding \texttt{MotorState}
struct, and then writes the computed pin levels. Storing state in the
\texttt{MotorState} array allows \texttt{GET\_STATUS} to report
current actuator conditions without querying hardware registers,
which is not possible on the ATmega328P once a pin has been
configured as output~\cite{atmel2015atmega328p}.

\subsubsection{Pump Control}

The pump motor is connected to a single-channel motor driver (Driver~3)
that exposes one speed (PWM) pin and one direction pin, differing from
the dual-IN-pin configuration used for the drive motors. The
\texttt{setPump} function applies the same validation and clamping
logic as \texttt{setMotor} and writes to \texttt{PUMP\_SPEED} and
\texttt{PUMP\_DIRECTION} accordingly.

\subsubsection{Safe Stop}

The \texttt{safeStop()} function sets the PWM speed pin of every
actuator to zero and drives all direction pins LOW. Writing zero to
the PWM register cuts motor current immediately; driving both direction
pins LOW simultaneously places the H-bridge in passive brake mode,
preventing back-EMF from the decelerating motors from feeding voltage
back into the direction pin outputs~\cite{horowitz2015art}. All
\texttt{MotorState} structs are reset to reflect the idle condition.
\texttt{safeStop()} is called unconditionally at startup, on any
unknown command, on buffer overflow, and on failsafe timeout, ensuring
that every error path leaves the hardware in a known safe state.

% ............................................................
\subsection{Safety Mechanisms}
\label{sec:safety}
% ............................................................

\subsubsection{Heartbeat and Failsafe Timeout}

The most critical safety feature of the firmware is its
\emph{communication watchdog}, implemented in \texttt{checkFailsafe()}.
The motivation for this mechanism is straightforward: if the Raspberry
Pi crashes, its software hangs, or the USB cable is disconnected while
motors are running, the Arduino has no way to receive a stop command.
Without a watchdog, the motors would continue running indefinitely —
a dangerous condition for a mobile robot carrying a pesticide
reservoir~\cite{kopetz2011real}.

The watchdog works as follows. Every time a valid command is received
and processed, \texttt{last\_command\_timestamp} is updated to
\texttt{millis()} — the Arduino's elapsed-time counter in
milliseconds. On every iteration of \texttt{loop()},
\texttt{checkFailsafe()} first checks whether any actuator is active
(speed $>$ 0). If all actuators are already idle, the check returns
immediately without resetting anything. If at least one actuator is
active, it evaluates:

\begin{equation}
\texttt{millis()} - \texttt{last\_command\_timestamp} > T_{\text{failsafe}}
\end{equation}

\noindent where $T_{\text{failsafe}} = 6000$\,ms is the configured
timeout. If the condition is true, \texttt{safeStop()} is called,
\texttt{ERROR:FAILSAFE\_STOP} is sent over the serial link, and the
timestamp is reset to prevent repeated triggering on subsequent
loop iterations.

The 6-second timeout is deliberately conservative: it is long enough
that normal command cadence (control messages sent multiple times per
second) never triggers a false positive, yet short enough that a
connection loss results in a stop within a time frame that prevents
significant unintended movement~\cite{kopetz2011real}.

\subsubsection{Buffer Overflow Protection}

The 64-byte command buffer is guarded against overflow: if an incoming
byte arrives when \texttt{command\_index} has reached the buffer limit,
the index is reset to zero, \texttt{safeStop()} is called, and
\texttt{ERROR:CMD\_TOO\_LONG} is returned. This prevents a
pathologically long or malformed transmission from corrupting the
buffer or advancing into adjacent memory — important on a
microcontroller with no memory protection unit
(MPU)~\cite{atmel2015atmega328p}.

\subsubsection{Unknown Command Handling}

Any command string that does not match a known verb causes
\texttt{safeStop()} to be called before the error response is sent.
This \emph{safe-by-default} policy means that a software error on the
Raspberry Pi that produces malformed output will halt the robot rather
than leaving it in an undefined state, consistent with the fail-fast
principle described in Section~\ref{sec:software}.

\bigskip
\noindent\textbf{[DIAGRAM 2 — Failsafe State Machine]}\\
\textit{A simple two-state diagram: \textbf{Active} (one or more
motors running) and \textbf{Idle} (all motors stopped). Transitions:
Active $\to$ Idle on \texttt{STOP\_ALL} command, valid command
timeout, unknown command, or buffer overflow; Idle $\to$ Active on
\texttt{SET\_MOTOR}/\texttt{SET\_PUMP} with speed $>$ 0. Annotate
the timeout transition with ``$\Delta t > 6000$\,ms $\Rightarrow$
\texttt{ERROR:FAILSAFE\_STOP}''. Place after this subsection.}
\bigskip

% ============================================================
%  Additional \bibitem entries
% ============================================================
%
% \bibitem{atmel2015atmega328p}
% Microchip Technology Inc.,
% ``ATmega328P 8-bit AVR Microcontroller with 32K Bytes In-System
% Programmable Flash Datasheet,''
% document no.\ DS40001906C, 2015.
%
% (All other entries — monk2016electronics, horowitz2015art,
%  kopetz2011real — are already listed in previous sections.)

 
% ------------------------------------------------------------
\section{Robot Software Implementation}
\label{sec:robot_software}
% ------------------------------------------------------------
 
This section documents the implementation of the Raspberry Pi software
stack introduced architecturally in Section~\ref{sec:software}. It
covers the organisation of the codebase, the rationale behind
structural decisions, and the implementation details of each major
component. The software is written entirely in Python~3.13 and follows
the layered, event-driven architecture described in
Chapter~3, with each architectural layer mapping directly to one or
more directories and modules in the project.
 
% ............................................................
\subsection{Project Structure}
\label{sec:project_structure}
% ............................................................
 
The project is organised as a single Python package named
\texttt{pest-detection-model-test}, managed with the \texttt{uv}
package manager and declared via a \texttt{pyproject.toml} build
manifest conforming to PEP~517~\cite{pep517}. Dependencies are pinned
in a \texttt{uv.lock} lockfile to ensure reproducible installation
across environments — a practice that proved essential when debugging
the runtime compatibility issues described in
Section~\ref{sec:dep_challenges}~\cite{pip2024}.
 
The top-level directory layout is as follows:
 
% \begin{verbatim}
% pest-detection-model-test/
% ├── app.py                  Main entry point
% ├── configs.py              Centralised configuration constants
% ├── config/                 Runtime-generated config files
% │   └── wifi_credentials.json
% ├── hardware/               Hardware abstraction layer
% │   ├── arduino.py          Serial communication wrapper
% │   └── motors.py           Motor control interface
% ├── services/               Service layer modules
% │   ├── asset_manager.py    Model file management
% │   ├── camera_service.py   Frame acquisition
% │   ├── health_service.py   Component health checks
% │   ├── inference_backend.py Inference API abstraction
% │   ├── model_service.py    Model loading and inference
% │   ├── network_service.py  WebSocket communication
% │   ├── watchdog_service.py Thread monitoring and recovery
% │   ├── wifi_service.py     Wi-Fi and hotspot management
% │   └── lifecycle_manager.py Shutdown callback registry
% ├── states/
% │   └── state_machine.py    FSM state definitions and transitions
% ├── utils/
% │   └── logger.py           Centralised logging setup
% ├── models/                 Downloaded model assets (.onnx / .tflite)
% ├── logs/                   Runtime log files
% └── output/                 Detection results and screenshots
% \end{verbatim}
 
The structure is a direct implementation of the five-layer software
architecture defined in Section~\ref{sec:software}. Table~\ref{tab:layer_map}
maps each architectural layer to its corresponding directory and
primary files in the codebase.
 
\begin{table}[h]
\centering
\caption{Mapping of architectural layers to codebase components.}
\label{tab:layer_map}
\begin{tabular}{p{3cm} p{3.5cm} p{5cm}}
\hline
\textbf{Architectural Layer} & \textbf{Directory / File} & \textbf{Responsibility} \\
\hline
Application Layer   & \texttt{app.py}                 & Entry point; boots, wires, and runs the system \\
State Management    & \texttt{states/}                & FSM state definitions and transition logic \\
Services Layer      & \texttt{services/}              & Camera, inference, networking, health, watchdog \\
Hardware Layer      & \texttt{hardware/}              & Arduino serial wrapper, motor interface \\
Utility Layer       & \texttt{utils/}, \texttt{configs.py} & Logging, configuration, shared helpers \\
\hline
\end{tabular}
\end{table}
 
The following subsections describe the role and implementation of each
component group in turn.
 
\subsubsection{Application Entry Point — \texttt{app.py}}
 
\texttt{app.py} is the sole entry point of the system, invoked with
\texttt{python app.py}. It is responsible for the application
lifecycle: initialising the logger, loading configuration from
\texttt{configs.py}, instantiating all hardware drivers and service
objects, injecting shared dependencies, and starting the main
execution loop. On shutdown — whether triggered by a keyboard
interrupt, a system command, or an unrecoverable fault — it calls the
\texttt{LifecycleManager} to invoke all registered cleanup callbacks
in reverse order, ensuring that the Arduino is commanded to idle and
the WebSocket connection is closed before the process
exits~\cite{gamma1994design}.
 
The entry point deliberately contains no business logic. Its only
purpose is to wire components together and hand control to the FSM.
This keeps the start-up sequence readable and ensures that each
individual service can be instantiated and tested in isolation without
invoking the full application~\cite{martin2008clean}.
 
\subsubsection{Centralised Configuration — \texttt{configs.py}}
 
All runtime parameters — serial port identifiers, baud rates, model
file paths, WebSocket endpoint URLs, inference thresholds, camera
resolution, and failsafe timeouts — are defined as typed constants in
\texttt{configs.py}. No magic numbers or hard-coded paths appear
elsewhere in the codebase; every module imports the values it needs
from this single file.
 
Centralised configuration serves two purposes. First, it provides a
single location to change when the system is deployed to different
hardware or a different network environment, reducing the risk of
inconsistent values scattered across multiple files~\cite{martin2008clean}.
Second, it makes the system's tunable parameters immediately visible
to anyone reading the code, without having to search through
individual modules. Sensitive values that vary between deployments
(Wi-Fi credentials, remote server addresses) are stored in
\texttt{config/wifi\_credentials.json}, which is generated at runtime
and excluded from version control via \texttt{.gitignore}.
 
\subsubsection{Hardware Layer — \texttt{hardware/}}
 
The \texttt{hardware/} directory contains two modules that implement
the Hardware Layer described in Section~\ref{sec:software}.
 
\texttt{arduino.py} wraps the \texttt{pyserial} library~\cite{pyserial}
to provide a high-level interface for sending commands to and
receiving responses from the Arduino over UART. It encapsulates
connection management (opening and closing the serial port), command
serialisation (encoding strings with a newline terminator), and
response parsing (reading until a newline and decoding the response
type). The rest of the codebase never interacts with \texttt{pyserial}
directly — all serial I/O flows through this wrapper, ensuring that
the underlying library can be swapped or mocked for testing without
modifying any other module~\cite{gamma1994design}.
 
\texttt{motors.py} sits above \texttt{arduino.py} and provides
a higher-level motor control interface. It translates semantic
commands — such as ``move forward at 60\% speed'' — into the
\texttt{SET\_MOTOR} command strings defined by the Arduino protocol
(Section~\ref{sec:serial_protocol}), and delegates their transmission
to \texttt{arduino.py}. This two-level abstraction mirrors the
Actuator Abstraction / Hardware Abstraction split in the Arduino
firmware: \texttt{motors.py} handles logical motor semantics while
\texttt{arduino.py} handles physical serial communication.
 
\subsubsection{State Machine — \texttt{states/state\_machine.py}}
 
The \texttt{state\_machine.py} module defines the \texttt{RobotState}
enumeration and the \texttt{StateMachine} class that governs system
behaviour. \texttt{RobotState} enumerates the seven system states
described in Section~\ref{sec:fsm}: \texttt{BOOTING},
\texttt{CHECKING}, \texttt{READY}, \texttt{STREAMING},
\texttt{DETECTING}, \texttt{ACTUATING}, and \texttt{ERROR}.
The \texttt{StateMachine} class maintains the current state, enforces
that only valid transitions are applied, and notifies registered
listeners when a transition occurs~\cite{hopcroft2006automata}.
 
Using a Python \texttt{Enum} for state values — rather than plain
string constants — means that an invalid state name produces an
\texttt{AttributeError} at the point of use rather than silently
propagating a typo through the system~\cite{martin2008clean}.
 
\subsubsection{Services Layer — \texttt{services/}}
 
The \texttt{services/} directory contains eight service modules, each
implementing one of the concurrent service threads described in
Section~\ref{sec:software}. Their roles are summarised in
Table~\ref{tab:services}.
 
\begin{table}[h]
\centering
\caption{Service modules and their responsibilities.}
\label{tab:services}
\begin{tabular}{p{4cm} p{8cm}}
\hline
\textbf{Module} & \textbf{Responsibility} \\
\hline
\texttt{asset\_manager.py}   & Verifies that model files exist in \texttt{models/}; downloads them from a remote URL if absent or corrupted. \\
\texttt{camera\_service.py}  & Initialises the \texttt{picamera2} backend, captures frames at the configured resolution, and places them on the shared frame queue. \\
\texttt{health\_service.py}  & Runs pre-flight checks on the camera, AI model, and Arduino serial link; reports component readiness to the FSM. \\
\texttt{inference\_backend.py} & Provides a unified inference API used by \texttt{model\_service.py}, abstracting over local (ONNX Runtime / TFLite) and remote inference providers. \\
\texttt{model\_service.py}   & Loads the detection model, dequeues frames from the camera, runs inference, and posts detection events to the FSM event queue. \\
\texttt{network\_service.py} & Manages the WebSocket client connection to the backend server; sends frames and telemetry upstream; receives and dispatches commands downstream. \\
\texttt{watchdog\_service.py} & Periodically checks that each service thread is alive and responsive; raises a fault event to the FSM if a thread becomes unresponsive. \\
\texttt{wifi\_service.py}    & Manages Wi-Fi association at startup and, if no network is available, provisions a local hotspot for device configuration. \\
\texttt{lifecycle\_manager.py} & Maintains a registry of cleanup callbacks; invoked by \texttt{app.py} on shutdown to tear down services in the correct order. \\
\hline
\end{tabular}
\end{table}
 
\noindent A key design decision common to all service modules is that
each service receives its dependencies — shared queues, the FSM
reference, hardware driver instances, configuration values — as
constructor arguments rather than importing global objects directly.
This dependency injection pattern~\cite{gamma1994design} means that
any service can be instantiated with mock dependencies in a test
environment without modifying the service code itself.
 
\subsubsection{Utility Layer — \texttt{utils/} and \texttt{configs.py}}
 
\texttt{logger.py} configures the Python standard \texttt{logging}
module with two handlers: a rotating file handler that writes
structured records to \texttt{logs/robot.log} and a console handler
that mirrors output to \texttt{stdout}. All modules obtain a logger
via \texttt{logging.getLogger(\_\_name\_\_)}, which automatically tags
every log record with the originating module name — essential for
tracing the source of events in a multi-threaded
system~\cite{martin2008clean}. Log rotation prevents unbounded disk
consumption on the Raspberry Pi's SD card during long deployments.
 
\subsubsection{Asset Management — \texttt{services/asset\_manager.py}}
 
The \texttt{asset\_manager.py} module addresses a practical
deployment concern: model files (\texttt{.onnx} or \texttt{.tflite})
are too large to store in the Git repository but are required before
inference can begin. The Asset Manager checks for the presence and
integrity of model files in the \texttt{models/} directory at startup.
If a file is absent or its checksum does not match the expected value,
the manager downloads it from a configured remote URL before the
system proceeds to the health check phase~\cite{warden2019tinyml}.
This eliminates the need for manual model deployment steps when
setting up the system on new hardware.
 
\bigskip
\noindent\textbf{[DIAGRAM 1 — Project Directory Structure and
Component Relationships]}\\
\textit{A dependency graph showing the six component groups as nodes
(\texttt{app.py}, \texttt{configs.py}, \texttt{hardware/},
\texttt{states/}, \texttt{services/}, \texttt{utils/}) with directed
arrows indicating import/dependency relationships. For example:
\texttt{app.py} $\to$ all others; every service $\to$
\texttt{configs.py} and \texttt{utils/logger.py};
\texttt{model\_service.py} $\to$ \texttt{inference\_backend.py};
\texttt{network\_service.py} $\to$ \texttt{hardware/arduino.py} via
\texttt{motors.py}. This gives a concise visual map of the codebase
without reproducing the full directory listing. Place after
Table~\ref{tab:services}.}
\bigskip
 
% ============================================================
%  Additional \bibitem entries
% ============================================================
%
% \bibitem{pep517}
% B.~Cannon et al., ``PEP 517 -- A build-system independent format
% for source trees,''
% \url{https://peps.python.org/pep-0517/}, 2015.
%
% \bibitem{pip2024}
% PyPA, ``pip -- The Python package installer,''
% \url{https://pip.pypa.io/}, 2024.
%
% \bibitem{pyserial}
% C.~Liechti, ``pyserial -- Python serial port access library,''
% \url{https://pyserial.readthedocs.io/}, 2020.
%
% (All other entries already listed in previous sections.)


% ............................................................
\subsection{State Machine Implementation}
\label{sec:fsm_impl}
% ............................................................
 
The Finite State Machine described architecturally in
Section~\ref{sec:fsm} is implemented in
\texttt{states/state\_machine.py} and comprises three tightly coupled
constructs: the \texttt{RobotState} enumeration, the
\texttt{StateTransition} dataclass, and the \texttt{StateMachine}
manager class. Together they provide a single, thread-safe authority
on what the robot is doing and what it is permitted to do at any given
moment~\cite{hopcroft2006automata}.
 
\subsubsection{State Definitions — \texttt{RobotState}}
 
Operational states are defined as a Python \texttt{Enum} rather than
plain string constants. Using an enumeration means that an invalid
state name raises an \texttt{AttributeError} at the call site rather
than silently propagating a typo into the transition
logic~\cite{martin2008clean}. The twelve states cover the full robot
lifecycle:
 
\begin{itemize}
    \item \textbf{Startup path:} \texttt{BOOTING} $\to$
          \texttt{WIFI\_CONNECTING} $\to$ \texttt{SERVER\_CONNECTING}
          $\to$ \texttt{CHECKING\_SYSTEMS} $\to$ \texttt{READY}.
    \item \textbf{Provisioning path:} \texttt{HOTSPOT\_MODE} $\to$
          \texttt{PROVISIONING} $\to$ \texttt{WIFI\_CONNECTING}
          (entered when no known network is available).
    \item \textbf{Operational states:} \texttt{REMOTE\_CONTROL},
          \texttt{AUTONOMOUS}, \texttt{DETECTING} (entered from
          \texttt{READY} once the system is fully initialised).
    \item \textbf{Fault and terminal states:} \texttt{ERROR} (with
          recovery paths back to \texttt{WIFI\_CONNECTING} or
          \texttt{CHECKING\_SYSTEMS}) and \texttt{SHUTDOWN}
          (terminal — no outgoing transitions).
\end{itemize}
 
\bigskip
\noindent\textbf{[DIAGRAM 1 — FSM State Transition Graph]}\\
\textit{A UML state diagram with all twelve states as rounded
rectangles. Use the Mermaid graph provided in the project
documentation as the authoritative source for transition arrows. Group
the startup path, provisioning path, operational states, and fault
states visually (e.g., using dashed boundary boxes or colour coding).
Mark \texttt{SHUTDOWN} as a terminal state with a filled circle. This
is the most important diagram in this subsection — place immediately
after this list.}
\bigskip
 
\subsubsection{Transition Validation}
 
The \texttt{StateMachine} class stores a
\texttt{VALID\_TRANSITIONS} dictionary that maps every source state to
the list of states reachable from it. Every call to
\texttt{transition\_to()} consults this graph before applying any
change: if the requested target is not in the allowed list, the method
logs a warning and returns \texttt{False}, leaving the current state
unchanged. This \emph{guard-and-reject} pattern ensures that no
component can drive the robot into an illegal state, regardless of
the order in which concurrent threads fire
events~\cite{hopcroft2006automata}.
 
On a successful transition, \texttt{transition\_to()} creates a
\texttt{StateTransition} dataclass instance recording the source
state, target state, reason string, timestamp, and an optional error
message. This record is appended to a capped history buffer
(default size: 100 entries) and all registered callbacks are invoked
outside the lock to prevent deadlocks. The method returns \texttt{True}
to the caller, which checks the return value before proceeding:
 
\begin{lstlisting}[language=Python,
                   caption={Guarded state transition in \texttt{app.py}.},
                   label={lst:transition}]
if not self.state_machine.transition_to(
    RobotState.READY,
    reason="All systems operational"
):
    
    self._rollback_startup()
    return False
\end{lstlisting}
 
% logger.error("Failed to enter READY – aborting startup") %
\subsubsection{Thread Safety}
 
The \texttt{StateMachine} class is accessed concurrently by multiple
threads: the main application loop, the watchdog, network callbacks,
and individual service threads. All reads and writes of the internal
state variable are protected by a \texttt{threading.Lock}
(\texttt{self.\_state\_lock}), which serialises access and prevents
race conditions~\cite{silberschatz2018os}. Callbacks are invoked
\emph{after} the lock is released so that a callback which itself
calls \texttt{get\_current\_state()} does not deadlock. Any exception
raised inside a callback is caught and logged; a buggy listener cannot
crash or stall the state machine.
 
\subsubsection{Public API and Usage Patterns}
 
The \texttt{StateMachine} exposes a minimal interface designed around
the principle that components should query system state rather than
track it themselves. Three query methods are particularly widely used:
 
\begin{itemize}
    \item \texttt{get\_current\_state()} — returns the current
          \texttt{RobotState}; used anywhere a component needs to
          branch on system status.
    \item \texttt{is\_ready()} — returns \texttt{True} when the
          state is \texttt{READY}, \texttt{REMOTE\_CONTROL}, or
          \texttt{AUTONOMOUS}; used as a gate before starting
          non-critical services.
    \item \texttt{can\_move()} — returns \texttt{True} in
          \texttt{REMOTE\_CONTROL}, \texttt{AUTONOMOUS}, or
          \texttt{DETECTING}; checked by the motor controller before
          accepting any movement command.
\end{itemize}
 
The callback registration mechanism (\texttt{register\_state\_change\_callback})
allows services to react to transitions without polling. For example,
the health service registers a callback that starts periodic checks
when the system enters \texttt{READY}, and the watchdog registers one
that initiates shutdown when \texttt{ERROR} is reached. This
event-driven coupling keeps services decoupled from one
another — they respond to FSM events without holding direct references
to each other~\cite{gamma1994design}.

% ............................................................
\subsection{Service Layer Implementation}
\label{sec:service_layer}
% ............................................................
 
The Services Layer forms the operational core of the Raspberry Pi
software stack. Each service encapsulates a single system concern —
image acquisition, AI inference, network communication, or health
monitoring — and runs in its own thread, interacting with other
services only through shared queues, the FSM event interface, or
explicit method calls on injected dependencies. This section documents
the implementation of each service in turn, beginning with the camera
subsystem. A recurring theme across all services is the use of
\emph{pluggable backend hierarchies} and \emph{fallback chains} that
allow the system to degrade gracefully when a preferred implementation
is unavailable — a design choice motivated directly by the runtime
compatibility challenges described in Section~\ref{sec:dep_challenges}.
 
% ............................................................
\subsubsection{Camera Service — \texttt{services/camera\_service.py}}
\label{sec:camera_service}
% ............................................................
 
The Camera Service is responsible for one task: continuously acquiring
frames from the camera hardware and making the most recent frame
available to other services with minimal latency. It exposes a
single-slot frame buffer rather than a queue, so that consumers always
receive the \emph{newest} image rather than waiting through a backlog
of stale frames — an important property for a real-time detection
pipeline where processing an outdated frame is worse than briefly
missing one~\cite{kopetz2011real}.
 
\paragraph{Pluggable Backend Architecture.}
 
A key design decision in the Camera Service is the separation of
\emph{what} the service does (capture and expose frames) from
\emph{how} it accesses the camera hardware. Camera APIs differ
significantly across platforms: the Raspberry Pi's \texttt{picamera2}
library uses the Linux \texttt{libcamera} stack and is not available
outside the Pi; OpenCV's \texttt{VideoCapture} is cross-platform but
uses a generic path that is less efficient on the Pi; and raw V4L2
access is Linux-only. Rather than hard-coding one API, the service
delegates all hardware interaction to a \emph{backend} class selected
at runtime from a preference-ordered list~\cite{gamma1994design}.
 
Three concrete backends are implemented, each in its own module under
\texttt{services/camera\_backends/}:
 
\begin{itemize}
    \item \textbf{\texttt{Picamera2Backend}} — uses the
          \texttt{picamera2} library to access the CSI-connected
          Raspberry Pi Camera Module via the \texttt{libcamera} ISP
          (Image Signal Processor). This is the preferred backend
          because it provides hardware-accelerated debayering,
          automatic exposure control, and lower CPU overhead than
          a generic video device~\cite{picamera2023}.
 
    \item \textbf{\texttt{OpenCVBackend}} — uses
          \texttt{cv2.VideoCapture} to open any V4L2-compatible
          camera device. Less efficient on the Pi but fully
          cross-platform, making it suitable for development and
          testing on a workstation without camera-specific
          drivers~\cite{bradski2000opencv}.
 
    \item \textbf{\texttt{V4L2CameraBackend}} — interfaces directly
          with the Linux Video4Linux2 kernel API, bypassing
          higher-level libraries entirely. Used as a last resort when
          neither \texttt{picamera2} nor OpenCV is installed.
\end{itemize}
 
\paragraph{Backend Selection and Caching.}
 
At startup, \texttt{CameraBackendFactory.select\_best\_backend()}
iterates over the preference list \texttt{["picamera2", "opencv",
"v4l2"]} in order. For each candidate it calls the backend's
\texttt{is\_available()} method, which checks whether the required
library is importable and the camera device is present. The first
backend that reports availability is instantiated. The selected result
is cached in \texttt{.camera\_backend\_cache.json} so that subsequent
startups skip the iteration and load the known-good backend
directly~\cite{gamma1994design}.
 
If the cached backend fails on a later startup — for example, because
the camera was disconnected — the factory falls back to iterating the
preference list again and updates the cache. This two-level strategy
avoids paying the discovery overhead on every launch without
sacrificing resilience.
 
\paragraph{Capture Loop.}
 
\texttt{CameraService.start()} spawns a background \emph{capture
thread} that runs \texttt{\_capture\_loop()} until a stop event is
set. The loop proceeds as follows on each iteration:
 
\begin{enumerate}
    \item \textbf{Backend open.} \texttt{\_open\_backend()} attempts
          to start the selected backend. On failure it retries with
          exponential back-off before raising a fault event to the
          FSM.
 
    \item \textbf{Frame read.} If the backend exposes a file
          descriptor (\texttt{fileno()}), \texttt{select.select()} is
          used with a configurable timeout to wait for data readiness
          without busy-polling. A longer timeout is applied on the
          first frame to allow the camera sensor time to settle
          exposure and white-balance~\cite{picamera2023}. Otherwise
          \texttt{backend.read()} is called directly.
 
    \item \textbf{Frame storage.} The captured frame is written into
          \texttt{\_latest\_frame} under a \texttt{threading.Lock},
          alongside a monotonically increasing \texttt{frame\_id} and
          a capture timestamp. The lock ensures a consumer reading
          \texttt{get\_latest()} never observes a partially written
          frame~\cite{silberschatz2018os}.
 
    \item \textbf{First-frame signal.} A \texttt{threading.Event} is
          set after the first successful frame write.
          \texttt{wait\_for\_first\_frame()} blocks on this event,
          allowing the startup sequence to confirm the camera is
          operational before the health check proceeds.
\end{enumerate}
 
\paragraph{Interface to Other Services.}
 
The Camera Service does not perform any frame encoding or network
transmission. It exposes two methods — \texttt{get\_latest()} and
\texttt{get\_latest\_metadata()} — returning the current frame and
its associated metadata (frame ID, timestamp, backend name)
respectively. The Model Service calls \texttt{get\_latest()} on each
inference cycle; the remote streaming path additionally JPEG-encodes
the retrieved frame before passing it to the Network Service. Keeping
the Camera Service unaware of both inference and networking means it
can be developed, tested, and replaced independently of either
downstream consumer~\cite{martin2008clean}.
 
\bigskip
\noindent\textbf{[DIAGRAM 1 — Camera Service Backend Selection and
Capture Flow]}\\
\textit{Two side-by-side sub-diagrams. Left: a flowchart of
\texttt{select\_best\_backend()} iterating the preference list,
calling \texttt{is\_available()}, branching to cache-and-return on
success or trying the next entry on failure, and reaching an ERROR
state if the list is exhausted. Right: the capture loop — open
backend, select/read frame, acquire lock, write
\texttt{\_latest\_frame} and metadata, release lock, set first-frame
event on first iteration, repeat. Place after this subsubsection.}
\bigskip


% ============================================================
%  Chapter 4 – Section 4.5.3 (continued)
%  AI Inference Service + Health Monitoring Service
% ============================================================

% ............................................................
\subsubsection{AI Inference Service — \texttt{services/model\_service.py}}
\label{sec:inference_service}
% ............................................................

The AI Inference Service is responsible for loading the object
detection model, running inference on incoming camera frames, and
dispatching detection results to the rest of the system. Like the
Camera Service, it abstracts hardware and runtime differences behind
a pluggable backend factory — a design that proved essential given
the runtime compatibility constraints documented in
Section~\ref{sec:dep_challenges_detail}.

\paragraph{Inference Backend Abstraction.}

All inference backends share a common abstract interface defined in
\texttt{services/inference\_backends/inference\_backend.py}. The
interface exposes three methods: \texttt{is\_available()}, which
checks that the required runtime library is importable;
\texttt{load\_model(path)}, which initialises the interpreter or
session from a model file; and \texttt{run(frame)}, which executes
a single forward pass and returns a list of detections. Two concrete
implementations are provided:

\begin{itemize}
	\item \textbf{\texttt{TFLiteBackend}} — wraps the
	\texttt{tflite\_runtime} interpreter. Preferred for
	on-device inference because TFLite is designed for
	low-memory ARM deployment and its interpreter has a smaller
	runtime footprint than ONNX Runtime~\cite{david2021tensorflow}.
	
	\item \textbf{\texttt{ONNXBackend}} — wraps an ONNX Runtime
	\texttt{InferenceSession}. Preferred on the server (x86-64)
	where ONNX Runtime wheels are readily available and the
	full RF-DETR operator set is
	supported~\cite{onnxruntime2023}.
\end{itemize}

\paragraph{Backend Selection and Fallback.}

\texttt{InferenceBackendFactory.select\_best\_backend()} receives a
\texttt{BackendConfig} specifying the preference order (e.g.,
\texttt{["tflite", "onnx"]}). For each candidate it first calls
\texttt{is\_available()}, then attempts \texttt{load\_model()} with
the configured model path. Both checks must pass before the backend
is accepted — a backend that installs correctly but fails to load
the specific model (due to unsupported operators, as experienced in
Section~\ref{sec:dep_challenges_detail}) is rejected and the next
candidate is tried. The first successfully loaded backend is cached
in \texttt{.backend\_cache.json} for subsequent startups.

If no local backend loads successfully, the service does not raise an
error immediately. Instead, \texttt{model\_service.is\_streaming()}
returns \texttt{False}, which \texttt{app.py} detects during startup
and uses as the trigger to activate the remote streaming fallback
path~\cite{warden2019tinyml}.

\paragraph{Streaming Modes.}

The Model Service supports two mutually exclusive inference modes,
reflecting the dual-path architecture described in
Section~\ref{sec:inference_pipeline}:

\begin{itemize}
	\item \textbf{Local streaming.} \texttt{start\_streaming()}
	launches a dedicated inference thread. On each iteration
	the thread calls \texttt{CameraService.get\_latest()} to
	retrieve the newest frame, passes it through
	\texttt{backend.run()}, enriches the raw detections with
	a timestamp and inference latency measurement, and
	dispatches the result to all registered detection listeners.
	The FSM is transitioned to \texttt{DETECTING} while
	inference is active.
	
	\item \textbf{Remote fallback streaming.}
	\texttt{start\_remote\_streaming()} launches a separate
	thread that retrieves frames from the Camera Service,
	JPEG-encodes them using OpenCV, and pushes the compressed
	bytes to the server via
	\texttt{NetworkService.update\_stream\_frame()}. The server
	executes inference and returns detection results through
	the WebSocket messaging channel. This path trades inference
	latency for the ability to run a model that exceeds the
	device's local capabilities.
\end{itemize}

The JPEG encoding step in the remote path is a deliberate compression
decision: sending raw frames over Wi-Fi at even modest resolutions
would saturate the available bandwidth, whereas JPEG at a quality
factor of 70--80\% reduces frame size by an order of magnitude with
minimal impact on detection accuracy for objects of the scale
targeted by this system~\cite{wallace1992jpeg}.

\paragraph{Detection Dispatch.}

Detection results — bounding boxes, class labels, and confidence
scores — are passed to all registered listener callbacks after each
inference cycle. This observer pattern~\cite{gamma1994design} keeps
the Model Service decoupled from the components that act on
detections (the FSM, the Network Service forwarding results to the
dashboard, and the actuation pipeline). Adding a new consumer of
detection events requires only registering a callback; no
modification to the Model Service is needed.

\bigskip
\noindent\textbf{[DIAGRAM 2 — Inference Service Backend Selection
	and Dual Streaming Modes]}\\
\textit{Left: a flowchart of \texttt{select\_best\_backend()} —
	iterate preference list, call \texttt{is\_available()}, call
	\texttt{load\_model()}, branch to cache-and-return on success or try
	next on failure; if list exhausted, set \texttt{is\_streaming =
		False} and trigger remote fallback. Right: two parallel swim lanes
	labelled ``Local Streaming'' and ``Remote Streaming'', showing the
	frame pull from Camera Service, the inference / encode step, and the
	output path (detection dispatch vs. NetworkService frame push). Place
	after this subsubsection.}
\bigskip

% ............................................................
\subsubsection{Health Monitoring Service —
	\texttt{services/health\_service.py}}
\label{sec:health_service}
% ............................................................

The Health Monitoring Service provides the system with a structured
view of the operational status of its critical components. It runs
two categories of checks: \emph{startup diagnostics}, which are
executed once during the initialisation sequence before the system
enters \texttt{READY}, and \emph{periodic checks}, which continue
during normal operation to detect degraded conditions that develop
over time.

\paragraph{Health Checks.}

Each check is an independent, callable unit that probes one system
component and returns a \texttt{HealthCheckResult} dataclass
containing the component name, a status code, a human-readable
message, and the time taken to complete the check. Four checks are
defined:

\begin{itemize}
	\item \textbf{Arduino connection} (\texttt{check\_arduino\_connection}).
	Sends a \texttt{PING} command over the serial link and
	verifies that the response is \texttt{PONG} within a
	timeout window. This confirms that the USB connection is
	active, the Arduino firmware is running, and the serial
	driver on the Pi is functional.
	
	\item \textbf{Motor health} (\texttt{check\_motors}).
	Sends a \texttt{CHECK} command and verifies the
	\texttt{MOTORS\_OK} response, confirming that the Arduino's
	motor controller is initialised and ready to accept
	actuation commands.
	
	\item \textbf{Camera availability} (\texttt{check\_camera}).
	Calls the injected \texttt{camera\_available\_fn} callable
	— normally \texttt{CameraService.is\_available()} — to
	confirm that the camera backend is open and has produced
	at least one frame. Using an injected callable rather than
	a direct reference to \texttt{CameraService} means the
	health check can be run in isolation with a mock function
	during testing~\cite{gamma1994design}.
	
	\item \textbf{AI model health} (\texttt{check\_ai\_model}).
	Calls the injected \texttt{model\_load\_fn} to attempt
	loading the inference backend. A failure here triggers the
	remote streaming fallback rather than an outright error,
	because the system can still operate — with reduced
	autonomy — without local inference capability.
\end{itemize}

\paragraph{Status Aggregation.}

Each \texttt{HealthCheckResult} carries one of four status values:
\texttt{OK}, \texttt{WARNING}, \texttt{ERROR}, or \texttt{SKIPPED}.
Results are stored in a dictionary keyed by component name.
\texttt{is\_all\_healthy()} returns \texttt{True} only if no result
carries an \texttt{ERROR} status; \texttt{get\_overall\_status()}
returns the highest-severity status across all results. This
distinction matters: a \texttt{WARNING} (for example, a slow model
load) does not block the startup sequence, whereas an \texttt{ERROR}
on the camera check causes an FSM transition to \texttt{ERROR} and
halts initialisation~\cite{kopetz2011real}.

\paragraph{Integration with the Startup Sequence.}

\texttt{app.py} calls the health service after the Camera Service
has confirmed its first frame and before the Model Service attempts
to load its backend. This ordering ensures that each check has a
meaningful subject to probe: the camera check requires the capture
thread to be running, and the model check requires the asset manager
to have verified model file availability. The result of the health
check phase determines which FSM state the system enters next:

\begin{itemize}
	\item All checks \texttt{OK} $\to$ \texttt{READY}.
	\item Camera or Arduino \texttt{ERROR} $\to$ \texttt{ERROR}
	(startup aborted).
	\item Model \texttt{ERROR} only $\to$ remote streaming activated,
	\texttt{READY} still reachable.
\end{itemize}

\paragraph{Watchdog Integration.}

During normal operation the watchdog service uses the health service
to inform recovery decisions. When the watchdog detects that a thread
has become unresponsive, it requests a fresh health check on the
affected component before deciding whether to attempt an in-process
restart or escalate to a full system reboot via the FSM
\texttt{SHUTDOWN} transition~\cite{kopetz2011real}.

\bigskip
\noindent\textbf{[DIAGRAM 3 — Health Service Check Flow and Startup
	Integration]}\\
\textit{A sequence diagram with four actors: \texttt{app.py},
	\texttt{HealthService}, individual check functions (Arduino, Motor,
	Camera, Model), and the FSM. Show \texttt{app.py} calling each check
	in order, the check returning a \texttt{HealthCheckResult}, and
	\texttt{app.py} evaluating \texttt{is\_all\_healthy()} before calling
	\texttt{state\_machine.transition\_to(READY)} or
	\texttt{transition\_to(ERROR)}. Annotate the model-error branch that
	leads to remote streaming activation rather than ERROR. Place after
	this subsubsection.}
\bigskip

% ============================================================
%  Service Interaction Summary
% ============================================================

\subsubsection{Service Interaction and Fallback Summary}
\label{sec:service_fallback}

The three services described above — Camera, Inference, and Health —
are tightly coordinated during the startup sequence but loosely
coupled during normal operation. Table~\ref{tab:fallback} summarises
the fallback behaviour when each service or its preferred
implementation fails.

\begin{table}[h]
	\centering
	\caption{Service failure modes and fallback behaviour.}
	\label{tab:fallback}
	\begin{tabular}{p{3.5cm} p{4cm} p{4.5cm}}
		\hline
		\textbf{Failure} & \textbf{Immediate Effect} & \textbf{Fallback / Resolution} \\
		\hline
		Camera backend unavailable (all three fail) &
		No frames produced; first-frame event never set &
		FSM transitions to \texttt{ERROR}; startup aborted \\
		Preferred camera backend fails (e.g. \texttt{picamera2} missing) &
		Factory iterates to next backend &
		\texttt{OpenCVBackend} or \texttt{V4L2CameraBackend} used \\
		Local inference backend fails (TFLite) &
		\texttt{is\_streaming()} returns \texttt{False} &
		Try \texttt{ONNXBackend}; if both fail, activate remote streaming \\
		Both local backends fail &
		No local inference &
		Remote streaming activated; frames sent to server for inference \\
		Remote streaming: server unreachable &
		Frame push fails; detection results unavailable &
		FSM transitions to \texttt{ERROR}; robot halted until network recovers \\
		Health check \texttt{ERROR} (camera or Arduino) &
		\texttt{is\_all\_healthy()} returns \texttt{False} &
		FSM transitions to \texttt{ERROR}; startup aborted \\
		Health check \texttt{ERROR} (model only) &
		Model check fails &
		Remote streaming fallback; \texttt{READY} still reachable \\
		\hline
	\end{tabular}
\end{table}

The consistent use of fallback chains across all three services
reflects the broader design philosophy of the system: prefer
graceful degradation over hard failure wherever the robot can still
provide useful function with reduced capability, but fail fast and
visibly when a component whose absence makes safe operation impossible
is unavailable~\cite{kopetz2011real,martin2008clean}.

% ============================================================
%  Additional \bibitem entries
% ============================================================
%
% \bibitem{wallace1992jpeg}
% G.~K.~Wallace, ``The JPEG still picture compression standard,''
% \textit{IEEE Transactions on Consumer Electronics},
% vol.~38, no.~1, pp.~xviii--xxxiv, 1992.
%
% (All other entries already listed in previous sections.)



% ============================================================
%  Chapter 4 – Section 4.5.4
%  Application Runtime and Lifecycle Management
% ============================================================

% ............................................................
\subsection{Application Runtime and Lifecycle Management}
\label{sec:lifecycle}
% ............................................................

A multi-threaded system that acquires external resources — serial
ports, camera devices, network sockets — must release those resources
reliably on exit, regardless of whether shutdown is triggered
normally, by an unhandled exception, or by an external signal. Failing
to do so can leave the Arduino in an active state, hold the camera
device open, or prevent the WebSocket connection from closing cleanly,
all of which require a physical power cycle to
recover~\cite{kopetz2011real}. The application addresses this through
a \texttt{LifecycleManager} component that decouples resource
acquisition from resource release, and through a clearly sequenced
startup and shutdown procedure orchestrated by the
\texttt{RobotApplication} class in \texttt{app.py}.

% ............................................................
\subsubsection{The LifecycleManager}
\label{sec:lifecycle_manager}
% ............................................................

The \texttt{LifecycleManager} implements a \emph{cleanup stack}: an
ordered list of \texttt{(name, cleanup\_fn)} pairs that grows during
startup as each subsystem registers its tear-down callback, and is
consumed in reverse order during shutdown. The reverse-order guarantee
is the central design property — it ensures that a service which
depends on another is always stopped before the service it depends
on~\cite{gamma1994design}.

The implementation is deliberately minimal:

\begin{lstlisting}[language=Python,
	caption={\texttt{LifecycleManager} — registration
		and reverse-order cleanup.},
	label={lst:lifecycle}]
	class LifecycleManager:
	def __init__(self):
	self._cleanup_stack = []
	self._lock = threading.Lock()
	
	def register(self, name: str, cleanup_fn) -> None:
	with self._lock:
	self._cleanup_stack.append((name, cleanup_fn))
	
	def cleanup_all(self) -> None:
	with self._lock:
	stack = list(reversed(self._cleanup_stack))
	self._cleanup_stack.clear()
	for name, fn in stack:
	try:
	fn()
	except Exception as exc:
	logger.exception("Cleanup failed for %s: %s", name, exc)
\end{lstlisting}

Two implementation details are worth noting. First, the \texttt{Lock}
protecting the stack guards against a race condition in which a
service starting in a background thread registers a callback at the
same moment that \texttt{cleanup\_all()} is
executing~\cite{silberschatz2018os}. Second, each cleanup invocation
is wrapped in a \texttt{try/except}: if one callback raises an
unhandled exception, the remaining callbacks still execute. A partial
cleanup — where only some resources are released — is safer than a
complete halt that leaves all remaining resources open.

% ............................................................
\subsubsection{Startup Sequence}
\label{sec:startup_seq}
% ............................................................

\texttt{RobotApplication.startup()} executes a strictly ordered
initialisation sequence. Each step is gated on the success of the
previous one: a failure at any point transitions the FSM to
\texttt{ERROR} and returns \texttt{False}, aborting startup cleanly
without entering the main loop. The sequence proceeds through five
phases:

\begin{enumerate}
	\item \textbf{Network phase.} The Wi-Fi service attempts to
	connect to a known network. On success the network service
	establishes the WebSocket connection to the backend server
	and registers \texttt{network\_service.disconnect} with the
	lifecycle manager. On failure the system enters
	\texttt{HOTSPOT\_MODE} and waits for provisioning before
	retrying.
	
	\item \textbf{Hardware phase.} The Arduino serial connection is
	opened and a \texttt{PING}/\texttt{PONG} handshake confirms
	firmware readiness. \texttt{arduino.disconnect} and
	\texttt{\_cleanup\_motors} are registered with the lifecycle
	manager. If the Arduino is unreachable, startup aborts —
	the system cannot actuate safely without a confirmed
	hardware link.
	
	\item \textbf{Camera phase.} The Camera Service is started and
	\texttt{wait\_for\_first\_frame()} blocks until the capture
	thread confirms the camera is producing frames.
	\texttt{camera\_service.stop} is registered once the first
	frame is received.
	
	\item \textbf{Health check phase.} The Health Service runs all
	four diagnostics (Section~\ref{sec:health_service}). A
	camera or Arduino \texttt{ERROR} aborts startup; a model
	\texttt{ERROR} activates the remote streaming fallback but
	allows startup to continue.
	
	\item \textbf{Model phase.} The Model Service attempts to load a
	local inference backend. If successful,
	\texttt{start\_streaming()} is called and
	\texttt{model\_service.stop\_streaming} is registered. If
	no local backend loads, \texttt{start\_remote\_streaming()}
	is called instead. The watchdog is then started and its
	cleanup callback registered last — ensuring it is the
	first to be stopped on shutdown.
\end{enumerate}

On completion of all five phases the FSM transitions to
\texttt{READY} and \texttt{startup()} returns \texttt{True}.
Table~\ref{tab:lifecycle_reg} lists the registration order and the
corresponding cleanup actions.

\begin{table}[h]
	\centering
	\caption{Lifecycle registration order and cleanup callbacks.}
	\label{tab:lifecycle_reg}
	\begin{tabular}{lll}
		\hline
		\textbf{Registration Order} & \textbf{Component} & \textbf{Cleanup Callback} \\
		\hline
		1 & NetworkService   & \texttt{network\_service.disconnect}  \\
		2 & Arduino          & \texttt{arduino.disconnect}           \\
		3 & Motors           & \texttt{\_cleanup\_motors}            \\
		4 & CameraService    & \texttt{camera\_service.stop}         \\
		5 & ModelService     & \texttt{model\_service.stop\_streaming} \\
		6 & WatchdogService  & \texttt{watchdog\_service.stop}       \\
		\hline
		\multicolumn{3}{l}{\textit{Shutdown executes in reverse: Watchdog
				$\to$ Model $\to$ Camera $\to$ Motors $\to$ Arduino $\to$ Network}} \\
		\hline
	\end{tabular}
\end{table}

% ............................................................
\subsubsection{Main Loop}
\label{sec:main_loop}
% ............................................................

After a successful startup, \texttt{RobotApplication.run()} enters
the main execution loop, which runs until the internal
\texttt{\_is\_running} flag is cleared. The loop is deliberately
lightweight: it does not perform camera capture, inference, or network
I/O directly — those tasks run in their own threads managed by the
respective services. The main loop's responsibilities are:

\begin{itemize}
	\item \textbf{Event processing.} Draining the FSM event queue
	and dispatching events (detection results, remote commands,
	health alerts) to the appropriate handlers.
	\item \textbf{Watchdog polling.} Checking that all monitored
	service threads are still alive and escalating to the FSM
	if any thread has become unresponsive.
	\item \textbf{Remote command handling.} Processing commands
	received from the server (movement, mode changes, shutdown
	requests) and translating them into FSM transitions or
	hardware commands.
\end{itemize}

Keeping the main loop thin — a dispatcher rather than a processor —
means that a slow handler in one iteration does not delay frame
capture or inference in the next, because those tasks are insulated
in their own threads~\cite{silberschatz2018os}.

% ............................................................
\subsubsection{Shutdown Sequence}
\label{sec:shutdown_seq}
% ............................................................

\texttt{RobotApplication.shutdown()} can be triggered by three
distinct sources: a POSIX signal (\texttt{SIGINT} / \texttt{SIGTERM})
delivered to the process, an \texttt{ERROR}$\to$\texttt{SHUTDOWN}
FSM transition escalated by the watchdog, or a remote shutdown
command received from the server. Regardless of trigger, the shutdown
procedure is identical:

\begin{enumerate}
	\item \textbf{Stop flag.} \texttt{\_is\_running} is set to
	\texttt{False}, causing the main loop to exit on its next
	iteration.
	\item \textbf{FSM transition.} The state machine is transitioned
	to \texttt{SHUTDOWN}, preventing any further state changes
	and signalling all callback listeners to cease activity.
	\item \textbf{Motor stop.} A \texttt{STOP\_ALL} command is sent
	to the Arduino immediately — before any service teardown
	— to ensure physical actuators are halted as early as
	possible in the shutdown sequence~\cite{kopetz2011real}.
	\item \textbf{Lifecycle cleanup.} \texttt{lifecycle\_manager.cleanup\_all()}
	is called, executing the registered callbacks in reverse
	order: Watchdog $\to$ Model $\to$ Camera $\to$ Motors
	$\to$ Arduino $\to$ Network.
\end{enumerate}

The explicit motor stop before \texttt{cleanup\_all()} is a safety
measure: if the Model or Camera service cleanup were to stall —
for example, due to a blocking thread join — the Arduino would
already be in an idle state rather than continuing to drive motors
during the delay.

\bigskip
\noindent\textbf{[DIAGRAM 1 — Application Lifecycle Sequence
	Diagram]}\\
\textit{A UML sequence diagram with five actors: \texttt{app.py},
	\texttt{LifecycleManager}, \texttt{FSM}, individual services (grouped
	as one actor for clarity), and \texttt{Arduino}. Show three
	horizontal phases separated by dashed lines: \textbf{Startup}
	(service init $\to$ register callback loop $\to$ FSM READY),
	\textbf{Main Loop} (event dispatch $\to$ watchdog poll, repeating),
	and \textbf{Shutdown} (stop flag $\to$ FSM SHUTDOWN $\to$ Arduino
	STOP\_ALL $\to$ cleanup\_all reverse stack). Annotate the
	registration arrows with the component name from
	Table~\ref{tab:lifecycle_reg}. This is the most important diagram
	in this subsection — place immediately after this paragraph.}
\bigskip


% ============================================================
%  Chapter 4 – Section 4.6
%  Server Implementation
% ============================================================

% ------------------------------------------------------------
\section{Server Implementation}
\label{sec:server}
% ------------------------------------------------------------

The backend server acts as the coordination hub between the robot and
the operator dashboard. It is implemented in Python using
\textbf{FastAPI}~\cite{fastapi2024}, an asynchronous web framework
that natively supports both HTTP and WebSocket endpoints within the
same application. The server exposes a REST API for robot registration
and command queuing, and a persistent WebSocket channel for real-time
bidirectional communication with the robot. All mutable runtime state
is held in two in-memory core services that are attached to the
FastAPI application instance at startup and accessed by route handlers
via dependency injection.

% ............................................................
\subsection{Architecture Overview}
\label{sec:server_arch}
% ............................................................

The server follows a three-layer structure:

\begin{itemize}
	\item \textbf{API Router} (\texttt{routers/control.py}) —
	declares all HTTP routes and the WebSocket endpoint,
	validates incoming payloads, and delegates business logic
	to the core services.
	
	\item \textbf{Core Services} (\texttt{core/robot\_store.py},
	\texttt{core/model\_adapters.py}) — \texttt{RobotStore}
	maintains an in-memory registry of connected robots and
	their status; \texttt{ModelRegistry} manages detection
	model metadata and adapter objects.
	
	\item \textbf{Dependency Injection Helpers} — lightweight
	functions (\texttt{get\_robot\_store},
	\texttt{get\_command\_queues},
	\texttt{get\_model\_registry}) that retrieve the shared
	service objects from \texttt{request.app.state}, keeping
	route function signatures clean and making unit testing
	straightforward~\cite{gamma1994design}.
\end{itemize}

The server is stateless from the HTTP perspective: no database is
used and all mutable state lives in the two core services. This
simplifies deployment — the server requires no external dependencies
beyond the Python runtime — but means that robot registrations and
queued commands are lost on process restart.

% ............................................................
\subsection{Data Models}
\label{sec:server_models}
% ............................................................

All request payloads are declared as Pydantic
\texttt{BaseModel} subclasses~\cite{pydantic2024}, which provide
automatic JSON schema validation and generate OpenAPI documentation
without additional code. Three models are defined:

\begin{itemize}
	\item \textbf{\texttt{RegisterRobotPayload}} (\texttt{robot\_id},
	\texttt{name}, \texttt{ip\_address}) — validates the
	registration request and supplies the fields needed to
	create a new \texttt{RobotStore} entry.
	
	\item \textbf{\texttt{RobotCommandPayload}} (\texttt{command})
	— validates that a command string is present before it is
	checked against the whitelist and enqueued.
	
	\item \textbf{\texttt{WifiPayload}} (\texttt{ssid},
	\texttt{password}) — carries Wi-Fi credentials to a robot
	in provisioning mode; only accepted when the robot's status
	field equals \texttt{"provisioning"}.
\end{itemize}

% ............................................................
\subsection{REST API}
\label{sec:server_rest}
% ............................................................

The REST API provides endpoints for robot lifecycle management and
model metadata retrieval. Table~\ref{tab:api_endpoints} lists all
endpoints, their methods, and their purpose.

\begin{table}[h]
	\centering
	\caption{Server REST API endpoints.}
	\label{tab:api_endpoints}
	\begin{tabular}{lp{5cm}p{5.5cm}}
		\hline
		\textbf{Method + Path} & \textbf{Handler} & \textbf{Description} \\
		\hline
		\texttt{POST /api/robots/register}
		& \texttt{register\_robot}
		& Creates a robot entry, initialises its command queue, returns a session token. \\
		\texttt{GET /api/robots}
		& \texttt{list\_robots}
		& Returns all registered robots and their current status. \\
		\texttt{GET /api/robots/\{id\}}
		& \texttt{get\_robot}
		& Returns a single robot; 404 if not found. \\
		\texttt{POST /api/robots/\{id\}/command}
		& \texttt{send\_command}
		& Validates against \texttt{VALID\_COMMANDS}, enqueues, returns \texttt{queued}. \\
		\texttt{POST /api/robots/\{id\}/wifi}
		& \texttt{send\_wifi\_credentials}
		& Queues Wi-Fi credentials; rejected if robot is not in provisioning mode. \\
		\texttt{GET /api/models}
		& \texttt{list\_models}
		& Returns names of all registered models. \\
		\texttt{GET /api/models/\{name\}}
		& \texttt{get\_model}
		& Returns model metadata (architecture, path). \\
		\texttt{GET /api/models/\{name\}/availability}
		& \texttt{check\_model\_availability}
		& Reports whether the model file exists on disk and is loaded. \\
		\hline
	\end{tabular}
\end{table}

\paragraph{Command Validation.}
The \texttt{send\_command} endpoint validates the incoming command
string against a \texttt{VALID\_COMMANDS} whitelist before enqueuing
it. Accepted commands are \texttt{start\_stream}, \texttt{stop\_stream},
\texttt{reboot}, and \texttt{stop}. Any string outside this set is
rejected with an HTTP 400 response. This whitelist prevents arbitrary
strings from reaching the robot's command handler and provides a clear
extension point: adding a new robot capability requires adding the
command to the list on both the server and the firmware.

\paragraph{Per-Robot Command Queue.}
Each registered robot is assigned a dedicated \texttt{asyncio.Queue}
at registration time. Commands from the HTTP endpoint are placed on
this queue asynchronously; the WebSocket handler consumes them and
forwards them to the robot. The queue guarantees FIFO ordering and
allows multiple dashboard clients to submit commands concurrently
without race conditions~\cite{tanenbaum2021networks}.

% ............................................................
\subsection{WebSocket Control Channel}
\label{sec:server_ws}
% ............................................................

The WebSocket endpoint at \texttt{/ws/control/\{robot\_id\}} provides
the persistent, bidirectional channel over which the robot and server
exchange real-time data. It is the server-side counterpart of the
\texttt{NetworkService} WebSocket client running on the Raspberry Pi
(Section~\ref{sec:service_layer}).

The endpoint manages three concurrent concerns within a single
\texttt{async} handler:

\paragraph{Command Delivery.}
A background \texttt{asyncio} task (\texttt{send\_commands}) runs
concurrently with the receive loop. It continuously pulls items from
the robot's command queue and pushes them as JSON messages over the
WebSocket. Using a background task rather than interleaving sends with
receives ensures that a slow or empty queue does not block the
heartbeat receive loop~\cite{fastapi2024}.

\paragraph{Heartbeat and Status Updates.}
The server expects a JSON message with \texttt{type: "heartbeat"}
from the robot at least once every \texttt{HEARTBEAT\_TIMEOUT\_SECONDS}
(40 seconds). The receive loop blocks on incoming messages with this
timeout; if no message arrives within the window, the socket is
closed and the robot is marked offline in \texttt{RobotStore}. On
each received heartbeat, \texttt{RobotStore.update\_status()} is
called via \texttt{asyncio.to\_thread()} to avoid blocking the
event loop with a potentially slow in-memory
operation~\cite{fastapi2024}.

\paragraph{Disconnection Handling.}
Both clean disconnects (\texttt{WebSocketDisconnect}) and unexpected
errors are caught in the handler. In either case the background
sender task is cancelled, any queued commands are discarded, and the
robot's status is updated to offline. This ensures the \texttt{RobotStore}
always reflects the actual connectivity state of each robot.

\bigskip
\noindent\textbf{[DIAGRAM 1 — Server WebSocket Lifecycle]}\\
\textit{A sequence diagram with three actors: Dashboard/Client,
	Server WebSocket Handler, and Robot (Pi). Show the three phases:
	Connect (\texttt{accept()} + background task spawn), Normal Operation
	(heartbeat loop on receive side; command queue drain on send side),
	and Disconnect (task cancel + RobotStore offline update). Annotate
	message types (\texttt{type: "heartbeat"}, JSON command frames) on
	each arrow. Place after this subsection.}
\bigskip

% ............................................................
\subsection{Validation and Error Handling}
\label{sec:server_validation}
% ............................................................

Three layers of validation protect the server from invalid input:

\begin{itemize}
	\item \textbf{Schema validation.} Pydantic models reject
	malformed or missing fields before a route handler is
	entered, returning an HTTP 422 response automatically with
	a structured error body~\cite{pydantic2024}.
	
	\item \textbf{Resource existence.} Routes that reference a
	\texttt{robot\_id} or \texttt{model\_name} raise an
	HTTP 404 \texttt{HTTPException} if the identifier is not
	found in the respective store.
	
	\item \textbf{State-dependent logic.} The Wi-Fi credential
	endpoint checks the robot's \texttt{status} field and
	rejects requests when the robot is not in provisioning
	mode, preventing credentials from being overwritten during
	normal operation.
\end{itemize}

% ............................................................
\subsection{Extensibility}
\label{sec:server_ext}
% ............................................................

The server architecture was designed with future extension in mind.
Three explicit extension points are identified:

\begin{itemize}
	\item \textbf{Authentication.} The registration and command
	endpoints include placeholder comments for token
	verification. Adding authentication requires implementing
	a verification function and inserting it as a FastAPI
	dependency on the relevant routes — no structural change
	to the router is needed.
	
	\item \textbf{Model hot-loading.} The \texttt{ModelRegistry}
	adapter pattern~\cite{gamma1994design} allows models to be
	loaded and unloaded at runtime without restarting the
	server or modifying the API. A new model adapter class is
	registered once; the existing \texttt{/api/models}
	endpoints expose it automatically.
	
	\item \textbf{Extended command set.} Adding a new robot
	capability requires appending its command string to
	\texttt{VALID\_COMMANDS} on the server and implementing
	the corresponding handler in the Arduino firmware or
	Raspberry Pi software. No changes to the WebSocket
	protocol or queue mechanism are needed.
\end{itemize}

% ============================================================
%  Additional \bibitem entries
% ============================================================
%
% \bibitem{fastapi2024}
% S.~Ramirez, ``FastAPI: Modern, fast web framework for building
% APIs with Python,''
% \url{https://fastapi.tiangolo.com/}, 2024.
%
% \bibitem{pydantic2024}
% S.~Colvin et al., ``Pydantic: Data validation using Python type
% hints,''
% \url{https://docs.pydantic.dev/}, 2024.